Towards completely caring 15-minute neighbourhoods

### 2025-01-30 paper/Manuscript.qmd
R1 submitted.

---
title: "Towards completely caring 15-minute neighbourhoods"
author:
  - name: Anastasia Soukhov 
    email: soukhoa@mcmaster.ca
    affiliations:
        - name: McMaster University
          department: School of Earth, Environment & Society
          city: Hamilton, Canada
    attributes:
        corresponding: true
  - name: Léa Ravensbergen
    email: ravensbl@mcmaster.ca
    affiliations:
        - name: McMaster University
          department: School of Earth, Environment & Society
          city: Hamilton, Canada
  - name: Lucía Mejía Dorantes
    email: lucia.mejia.dorantes@gmail.com
    affiliations:
        - name: Consultant
          city: Karlsruhe, Germany
  - name: Antonio Páez
    email: paezha@mcmaster.ca
    affiliations:
        - name: McMaster University
          department: School of Earth, Environment & Society
          city: Hamilton, Canada
format: 
  # docx
  elsevier-pdf:
    keep-tex: true
    include-in-header:
      text: |
        \usepackage{graphicx}
        \usepackage{unicode-math}
        \usepackage{hyperref}
        \def\tightlist{}
        \usepackage{setspace}
        \doublespacing
        \usepackage{lineno}
        \linenumbers
    journal:
      #name: "Transportation Research Part A: Policy and Practice"
      formatting: preprint
      model: 3p
      cite-style: authoryear
keywords: ["Accessibility", "Mobility of Care", "Gender", "15-minute neighbourhood", "Care Complete Neighbourhoods"]
bibliography: bibliography.bib
---

@@

@@

The "15-Minute City" concept has been embraced by global leaders to promote human-scale neighbourhoods with transport and land-use designs that support short trips to daily necessities. This paper bridges the 15-Minute City to "Mobility of Care",  a framework that foregrounds travel to care destinations, travel done predominately by women. This focus contrasts the more commonly studied travel to employment and leisure destinations. While the 15-Minute City concept is flexible enough to consider all destination types, gendered examinations are relatively lacking in the literature, and little research to date has focused explicitly on care destinations. This gap is addressed in this paper by identifying which areas in the city of Hamilton, Canada are ‘caring 15-minute neighbourhoods’. To do so, a database of care destinations was compiled to estimate the number (using the cumulative opportunity accessibility measure) and diversity of mix (using the entropy measure) of care destinations within a 15-minute walk from residential parcels. Using data-driven machine learning techniques (Self-Organizing Maps and Decision Trees), neighborhoods were classified into 'caring 15-minute neighbourhood' typologies that are examined across residential socio-economic profiles. Our results suggest that the majority of caring 15-minute neighbourhoods are in the urban core, where more lower-income households reside. In contrast, not caring 15-minute neighbourhoods in higher-income peripheral areas. Areas that make good candidates for urban policy intervention are identified, and the implications of enhancing 15-minute walkable caring access are discussed in relation to gender mainstreaming in transportation planning.

@@

From another perspective, feminist geographers, urban planners and other researchers agree that urban structure influences travel, but city planning falls short in planning for a neutral identity. In contrast, cities ought to be planned according to the multiple identities of the inhabitants [@vacchelli_towards_2018; @urbandevelopmentviennaGenderMainstreamingUrban2013]. Researchers have demonstrated how mobility patterns differ according to gender [@lawWomenTransportNew1999; @cresswell_gendered_2008; @levy2013travel; @little_gender_1994; @tronto_toward_1990; @neto_gender_2015]. The causes for this are varied, and related to gendered social norms and cultural factors that play a role in how, where, with whom, what for, and how far we move. Women tend to travel shorter distances and more frequently [@roorda_triap_2010; @morency_distance_2011], and conduct more activities related to care no matter their stage of life [@ilo_care_2018; @garciaromanGenderDifferencesTime2022]. Combined, these findings indicate that the spatial organization of activities has a different impact according to gender, and therefore the 15-Minute City concept should not be gender blind.

This work builds a theoretical bridge between the 15-Minute City and Mobility of Care concepts to answer questions about what destinations matter and who benefits from proximity planning. The Mobility of Care term was coined in @sanchezdemadariagaMobilityCareIntroducing2013; it refers to all travel required to sustain the needs of a household, such as grocery shopping, escorting children, travelling to health appointments, and running errands.  While decades of research have examined types of household-serving trips (e.g., shopping trips, escorting to school), @sanchezdemadariagaMobilityCareIntroducing2013 was the first to consider them all as one category. In doing so, care-related mobility can be shown to comprise a significant proportion of daily travel, characterizing approximately 30% of adults’ daily trips [@sanchezdemadariagaMobilityCareIntroducing2013; @sanchezdemadariagaMeasuringMobilitiesCare2019; @ravensbergen2023exploratory]. We argue that Mobility of Care is especially relevant to the 15-Minute City, as care trips are necessary for _all_ people and they most often occur at the local level. Moreover, care trips are of particular gendered and socio-economic importance, as women, and especially those within lower income households, complete the majority of care trips [@ravensbergen2023exploratory]. Despite care travel's importance in advancing gender equality -and equity more broadly-, Mobility of Care and access to care destinations have been under-examined relative to employment destinations. This is especially pertinent in accessibility research that has largely focused on employment points of interest e.g., [@farberOntarioLineSocioeconomic2019; @duarteInfluenceJobAccessibility2023; @ryanAccessibilitySpaceTime2023].

@@

# Data

@@

This work focuses on Hamilton, Ontario, a mid-sized city on the shore of Lake Ontario. Hamilton has a heterogenous land-use, with a populated and dense urban core, surrounded by suburban development, which is itself surrounded by rural communities. The Niagara Escarpment runs through Hamilton, and results in a city with two key elevations: a more dense lower city that contains the downtown core and the elevated suburban development referred to as ‘the Mountain’. In this work, we analyse the residential parcel centroids, 143,882 locations. We aggregated the points at the level of Canadian Census Dissemination Area (DA) along with the population and population per parcel plots in @fig-Fig1.

@@

\newpage

@@

flextable(dat) %>%
  ftExtra::colformat_md() %>%
  compose(
    i = 1, j = c(1,2,3), part = "header", value = as_paragraph(c("Care category", "Care destination types", "Sources"))) %>%
  width(j = 1, width = 1.7, unit = "cm") %>%
  width(j = 2, width = 10, unit = "cm") %>%
  width(j = 3, width = 5.3, unit = "cm")|>
  flextable::line_spacing(space = 0,
                          part = "body")
```

@@

## Accessibility: the cumulative opportunity measure

To capture the quantity of access to each type of destinations, a cumulative opportunity accessibility score $S^t_i$ is calculated. Scores for each of the 14 care destination types $t$ is calculated for every parcel $i$. The calculation takes the following mathematical form:

$$
S^t_i=\sum_{j}O_j^{t}\cdot f(c_{ij})
$$
\noindent Where:

- $i$ is a set of parcel point origin locations.
- $j$ is a set of care destination locations of type $t$.
- $O^t_j$ is a number of opportunities of category type $t$ at $j$.
- $c_{ij}$ is the travel cost between $i$ and $j$.
- $f(\cdot)$ is an impedance function of $c_{ij}$; within the cumulative opportunity approach, it is a binary function that takes the value of 1 if $c_{ij}$ is less than a selected value [@handyMeasuringAccessibilityExploration1997].
- $S_{i}^t$ is the cumulative opportunity accessibility score, the sum of weighted opportunities reachable within $f(\cdot)$, at each $i$ for each $t$.

@@

$$
D_{i} = \frac{-\sum_{t} [S_i^{t} / \sum_{t}S^{t}_{i} \times ln(S_i^{t} / \sum_{t}S^{t}_{i} )]}{ln(n_t)}
$$

@@

- $i$ is a set of parcel point origin locations.
- $t$ is a set of care destination types (e.g., school, grocery, park, etc.)
- $n_t$ is the count of care destination types $t$. In this work, this value is 14.
- $S_{i}^t$ is the cumulative opportunity accessibility score, the sum of weighted opportunities reachable within a 15-minute walk from $i$.
- $D_i$ is the diversity score.

@@

## Overview of access to care destinations
How many care destinations can be reached within a 15-minute walk is summed for each of the four care categories defined in @tbl-Tbl1. The median parcel value for each DA is visualised in @fig-Fig4 and the median parcel diversity of care destinations accessible is presented in @fig-Fig5. Together, @fig-Fig4 and @fig-Fig5 represent summaries of 15 variables that served as inputs into the SOM algorithm.

```{r plot-figure4, out.width=600, out.height=600}
#| label: fig-Fig4
#| fig-cap: "The number of care destinations that can be reached within a 15-min walk per care category for a median parcel in each DA. The values are a summary of the 14 acessibility scores that served as inputs into the SOM. Basemap shapefiles are retrieved from the 2021 Canadian census [@governmentofcanadaCensusPopulation2023], the Open Data Hamilton Portal [@opendatahamiltonCityBoundary2023] and the USGS [@greatlakesUSGS2010]."
knitr::include_graphics(glue::glue(here::here(), "/figures/percats_copp_plot.png"))
```

```{r plot-figure5, out.width=450, out.height=450}
#| label: fig-Fig5
#| fig-cap: "The diversity measure based on the proportion of care category accessibility in Figure 4. These values are also a summary of the 15th input into the SOM. Basemap shapefiles are retrieved from the 2021 Canadian census [@governmentofcanadaCensusPopulation2023], the Open Data Hamilton Portal [@opendatahamiltonCityBoundary2023] and the USGS [@greatlakesUSGS2010]."
knitr::include_graphics(glue::glue(here::here(), "/figures/all_cats_entropy_copp_plot.png"))
```

\newpage

@@

In @fig-Fig4, the concentration of opportunities for all category types is centered in the downtown of Hamilton, with the highest concentration being in the lower city (near the lake shore) and pockets of the high concentration further from the shoreline. Grocery-centric destinations appear to be the most concentrated, followed by health-centric and dependent-centric caring destinations. Errand-centric destinations are the most sprawled. In many ways, @fig-Fig4 mirrors the spatial distribution of care destinations (@fig-Fig3) as the range of 15-minutes of travel is small. In @fig-Fig5, areas that have high care accessibility tend to have high diversity as well, though there are exceptions in pockets of the city outside the downtown core that have higher diversity but low levels of accessibility for all care categories. Similarly, there are areas with low diversity within the downtown core that have only moderate or high accessibility to certain care destination types.

Based on the SOM methodology discussed, 7 superclusters are identified from the 10 by 10 grid of SOM nodes. As diagnostics for the selection of the number of superclusters, the dissimilarity-index-based dendrogram with the proportion of parcels represented in each supercluster alongside the variance explained plot is visualised in @fig-Fig6. Labels representing grades A to D qualifying the 7 superclusters are then assigned. These grades as also reflected in @fig-Fig6 and represent the quantity of accessibility and diversity in accessibility per care category. Higher grades (A+ and A) corresponding to the highest accessibility and diversity, while lowest grades (D) representing lowest accessibility and diversity scores.

```{r plot-figure6, out.width = '100%', out.height = '100%'}
#| label: fig-Fig6
#| fig-cap: "The 7 resulting superclusters from the SOM the 10 by 10 grid output represented in a dendrogram (left) and the proportion of variance unexplained by cluster (right)."
knitr::include_graphics(glue::glue(here::here(), "/figures/all_access_sc_plot_small.png"))
```

\newpage

The grade for each supercluster was carefully assigned by the authors by referring to descriptive statistics and their notable trends. These statistics are summarised in @tbl-Tbl2 and in the boxplots in @fig-Fig7.

- A+ and A have exceptionally high caring accessibility for all destination types and high diversity scores in the top quantile or above. Together, these grades represent 24% of all parcels in the city.
- A- has high caring accessibility, but low diversity. Interestingly, A- is like A but with much higher dependent-centric destinations, particularly parks, schools, and daycares, with only moderately high scores for all other destination types. This disbalance lowers its overall diversity score. However, it can be characterized as a supercluster that demonstrates potential in being retrofitted to provide A+ or A level of completely caring access. 4% of all parcels are represented by the A- grade. 
- B+ and B superclusters present about average completely caring access. These superclusters can serve as the benchmark for what 'average' completely caring access in Hamilton currently looks like. These grades represent 15% of all parcels.
- B- provides above average dependent-centric destination access, particularly parks, daycares and schools, but below average access to other destination types and hence has low diversity scores. B- is like A- as it demonstrates complete caring 15-minute potential if retrofitted. As these parcels demonstrate caring access to some destinations, they may have the potential to be retrofitted to support complete access to all caring destination types. This grade represents 19% of all parcels.
- D superclusters demonstrate the lowest scores all-around, representing room for land-use improvement that addresses complete and caring 15-minute access. This supercluster characterizes the largest number of parcels, representing 38% of parcels in the city.

@@

\newpage

To spatially demonstrate where the superclusters are located, representative supercluster labels for each DA are visualised in @fig-Fig8. This visualisation is created by grouping parcels by their DA, calculating the median supercluster label for each label, and selecting the label that is most dominant within that DA. For reference, the median number of parcels in each DA is `r df_DA$parcel_count |> median() |> round()` and supercluster label membership within a DA is typically pure, with the median membership being `r df_DA$max_label_val |> median() |> scales::percent()` of a single cluster.

```{r plot-figure8, out.width=450, out.height=450}
#| label: fig-Fig8
#| fig-cap: "The maximum median parcel supercluster membership per DA. Escarpment is visualised as a grey dashed line. Basemap shapefiles are retrieved from the 2021 Canadian census [@governmentofcanadaCensusPopulation2023], the Open Data Hamilton Portal [@opendatahamiltonCityBoundary2023] and the USGS [@greatlakesUSGS2010]."
knitr::include_graphics(glue::glue(here::here(), "/figures/SClabel_plot.png"))
```

The resulting superclusters visualised in @fig-Fig8 are a combination of the cumulative opportunity and diversity of access measures translated into meaningful typologies through the SOM methodology. To reiterate, accessibility is the measure of _potential_ interaction, e.g., how many care destinations one could reach within a 15-minute walk, these typologies are useful in identifying which areas of the city are providing regionally high levels of complete caring access, which areas are regionally average, and which areas are regionally below average. Furthermore, neighbourhoods which provide disbalanced access can be further investigated.

In @fig-Fig8, it can be observed that the excellent and exceptional neighbourhoods (A, A+), which are very caring and very complete 15-minute neighbourhoods, are located within the center of the city and closest to the shoreline in the downtown core of Hamilton. The above and below average neighbourhoods (B+ B) are often proximate to the excellent and exceptional neighbourhoods and are more prevalent in south of the escarpment within the center Hamilton. The escarpment is a physical barrier, with few pedestrian-accessible access points to traverse; hence the typologies describing neighbouring DAs separated by the escarpment are often different. Below average (D) neighbourhoods are located in peripheral areas outside the center of Hamilton, in areas where urban form is characterised by lower density residential housing and auto-dependent mobility.

>>>>

---
title: "Towards completely caring 15-minute neighbourhoods"
author:
  - name: Anastasia Soukhov 
    email: soukhoa@mcmaster.ca
    affiliations:
        - name: McMaster University
          department: School of Earth, Environment & Society
          city: Hamilton, Canada
    attributes:
        corresponding: true
  - name: Léa Ravensbergen
    email: ravensbl@mcmaster.ca
    affiliations:
        - name: McMaster University
          department: School of Earth, Environment & Society
          city: Hamilton, Canada
  - name: Lucía Mejía Dorantes
    email: lucia.mejia.dorantes@gmail.com
    affiliations:
        - name: Consultant
          city: Karlsruhe, Germany
  - name: Antonio Páez
    email: paezha@mcmaster.ca
    affiliations:
        - name: McMaster University
          department: School of Earth, Environment & Society
          city: Hamilton, Canada
format: 
  # docx
  elsevier-pdf:
    keep-tex: true
    include-in-header:
      text: |
        \usepackage{graphicx}
        \usepackage{unicode-math}
        \usepackage{hyperref}
        \def\tightlist{}
        \usepackage{setspace}
        \doublespacing
        \usepackage{lineno}
        \linenumbers
        \floatplacement{figure}{H}
    journal:
      #name: "Transportation Research Part A: Policy and Practice"
      formatting: preprint
      model: 3p
      cite-style: authoryear
keywords: ["15-Minute City", "Chrono-Urbanism","Accessibility", "Mobility of Care", "Inequality", "Gender Gap"]
bibliography: bibliography.bib
---

@@

```{r to calc perc urbanised parcels in each scluster}
load(glue::glue(here::here(), "/data/data-raw/parcels/R_PARCELS_2020.rda"))
R_PARCELS_2020 <- R_PARCELS_2020 |> select(c("ID","URBAN_RURA")) |> st_drop_geometry()
R_PARCELS_2020$ID <- as.character(R_PARCELS_2020$ID)

access_15_urban_summary <- access_15_wide |> 
  select(c("from_id","scluster")) |>
  left_join(R_PARCELS_2020,by=c("from_id"= "ID"))

access_15_urban_summary <- access_15_urban_summary |> group_by(scluster,URBAN_RURA) |> 
  summarise(count_urb_parcels = n(), 
            .groups = "drop")
```

@@

The "15-Minute City" concept has been embraced by global leaders to promote human-scale neighbourhoods with transport and land-use designs that support short trips to daily necessities. This paper bridges the 15-Minute City to "Mobility of Care",  a framework that foregrounds travel to care destinations, travel done predominately by women. This focus contrasts the more commonly studied travel to employment and leisure destinations. While the 15-Minute City concept is flexible enough to consider all destination types, gendered examinations are relatively lacking in the literature, and little research to date has focused explicitly on care destinations. This gap is addressed in this paper by identifying which areas in the city of Hamilton, Canada are ‘caring 15-minute neighbourhoods’. To do so, a database of care destinations was compiled to estimate the number (using the cumulative opportunity accessibility measure) and diversity of mix (using the entropy measure) of care destinations within a 15-minute walk from residential parcels. Using data-driven machine learning techniques (Self-Organizing Maps and Decision Trees), neighborhoods were classified into 'caring 15-minute neighbourhood' typologies that are examined across residential socio-economic profiles. Our results suggest that the majority of caring 15-minute neighbourhoods are in the urban core, where more lower-income households reside. In contrast, not caring 15-minute neighbourhoods in higher-income peripheral areas. Areas that make good candidates for urban policy intervention are identified and the implications of enhancing 15-minute walkable caring access are discussed in relation to gender mainstreaming in transportation planning and limitations of this work.

@@

An increasing number of human geographers, planners and other researchers agree that urban structure influences travel, but city planning falls short in planning for a neutral identity. In contrast, cities ought to be planned according to the multiple identities of the inhabitants [@vacchelli_towards_2018; @urbandevelopmentviennaGenderMainstreamingUrban2013]. Researchers have demonstrated how mobility patterns differ according to gender [@lawWomenTransportNew1999; @cresswell_gendered_2008; @levy2013travel; @little_gender_1994; @tronto_toward_1990]. The causes for this are varied, and related to gendered social norms and cultural factors that play a role in how, where, with whom, what for, and how far we move. Women tend to travel shorter distances and more frequently [@roorda_triap_2010; @morency_distance_2011], and conduct more activities related to care no matter their stage of life [@ilo_care_2018; @garciaromanGenderDifferencesTime2022]. Combined, these findings indicate that the spatial organization of activities has a different impact according to gender, and therefore the 15-Minute City concept should not be gender blind.

This work builds a theoretical bridge between the 15-Minute City and Mobility of Care concepts to answer questions about what destinations matter and who benefits from proximity planning. The Mobility of Care term was coined in @sanchezdemadariagaMobilityCareIntroducing2013; it refers to all travel required to sustain the needs of a household, such as grocery shopping, escorting children, travelling to health appointments, and running errands.  While decades of research have examined types of household-serving trips (e.g., shopping trips, escorting to school), @sanchezdemadariagaMobilityCareIntroducing2013 was the first to consider them all as one category. In doing so, care-related mobility can be shown to comprise a significant proportion of daily travel, characterizing approximately 30% of adults’ daily trips [@sanchezdemadariagaMobilityCareIntroducing2013; @sanchezdemadariagaMeasuringMobilitiesCare2019; @ravensbergen2023exploratory]. We argue that Mobility of Care is especially relevant to the 15-Minute City, as care trips are necessary for _all_ people and they most often occur at the local level. Moreover, care trips are of particular gendered and socio-economic importance, as women, and especially those within lower income households, complete the majority of care trips [@ravensbergen2023exploratory]. Despite care travel's importance in advancing gender equality -and equity^[The United Nations defines "gender equality" as equal rights, responsibilities, and opportunities for all genders, regardless of gender at birth [@unwomenHandbookGenderMainstreaming2022]; they prefer "gender equality," as "equity" relates to the concept of fairness that are usually based on traditions, customs, or cultural beliefs which can perpetuate disadvantage. However, "equity," in the broader context of social justice, is widely accepted. As defined in the UNICEF Glossary of DEI terms, equity is the "process of being fair to all individuals and groups, by addressing present and historical inequality in order to work towards equality in outcomes" and may involve temporary special measures to address systemic discrimination faced by marginalized groups [@unicefAnnexDIVERSITYEQUITY2024].] more broadly-, Mobility of Care and access to care destinations have been under-examined relative to employment destinations. This is especially pertinent in accessibility research that has largely focused on employment points of interest e.g., [@farberOntarioLineSocioeconomic2019; @duarteInfluenceJobAccessibility2023; @ryanAccessibilitySpaceTime2023].

@@

# Case study and data

@@

This work focuses on Hamilton, Ontario, a mid-sized city on the shore of Lake Ontario. Hamilton has a heterogenous land-use, with a populated and dense urban core, surrounded by suburban development, which is itself surrounded by rural communities. The Niagara Escarpment runs through Hamilton, and results in a city with two key elevations: a more dense lower city that contains the downtown core and the elevated suburban development referred to as ‘the Mountain’. In this work, we analyse the residential parcel centroids, 143,882 locations in the city [@Teranet_2021]. We aggregated the points at the level of Canadian Census Dissemination Area (DA) along with the population and population per parcel plots in @fig-Fig1. The DA is the most spatially granual level of publicly available Canadian census data. DAs are created by Statistics Canada for the purpose of the census and each represent between 400 to 700 people (Q1:442 and Q3:664 from @fig-Fig1).

@@

@@

flextable(dat) %>%
  ftExtra::colformat_md() %>%
  compose(
    i = 1, j = c(1,2,3), part = "header", value = as_paragraph(c("Care category", "Care destination types", "Sources"))) %>%
  width(j = 1, width = 3, unit = "cm") %>%
  width(j = 2, width = 9, unit = "cm") %>%
  width(j = 3, width = 5, unit = "cm")|>
  flextable::line_spacing(space = 0,
                          part = "body")

```

@@

## The cumulative opportunity measure of accessibility

To capture the quantity of spatial access to each type of destinations, a cumulative opportunity accessibility score $S^t_i$ is calculated. Scores for each of the 14 care destination types $t$ is calculated for every parcel $i$. The calculation takes the mathematical form in @eq-access-cumulative:

$$
S^t_i=\sum_{j}O_j^{t}\cdot f(c_{ij})
$${#eq-access-cumulative}
\noindent Where:

- $i$ is a set of parcel point origin locations.
- $j$ is a set of care destination locations of type $t$.
- $O^t_j$ is a number of opportunities of category type $t$ at $j$.
- $c_{ij}$ is the travel cost between $i$ and $j$.
- $f(\cdot)$ is an impedance function of $c_{ij}$; within the cumulative opportunity approach, it is a binary function that takes the value of 1 if $c_{ij}$ is less than a selected value [@handyMeasuringAccessibilityExploration1997].
- $S_{i}^t$ is the cumulative opportunity accessibility score, the sum of opportunities reachable within $f(\cdot)$, at each $i$ for each $t$.

@@

$$
D_{i} = \frac{-\sum_{t} [S_i^{t} / \sum_{t}S^{t}_{i} \times ln(S_i^{t} / \sum_{t}S^{t}_{i} )]}{ln(n_t)}
$${#eq-entropy-measure}

@@

- $i$ is a set of parcel point origin locations.
- $t$ is a set of care destination types (e.g., school, grocery, park, etc.)
- $n_t$ is the count of care destination types $t$. In this work, this value is 14.
- $S_{i}^t$ is the cumulative opportunity accessibility score, the sum of opportunities reachable within a 15-minute walk from $i$.
- $D_i$ is the diversity score.

@@

## Overview of access to care destinations
The number of care destinations that can be reached within a 15-minute walk is summarised for each of the four care categories (as defined in @tbl-Tbl1). The median parcel value for each DA is visualised in @fig-Fig4 and the median parcel diversity of care destinations accessible is presented in @fig-Fig5. Together, @fig-Fig4 and @fig-Fig5 represent summaries of 15 variables were inputs into the SOM algorithm.

```{r plot-figure4, out.width=600, out.height=600}
#| label: fig-Fig4
#| fig-cap: "The number of care destinations that can be reached within a 15-min walk per care category for a median parcel in each DA. The values are a summary of the 14 acessibility scores that are inputs into the SOM. Basemap shapefiles are retrieved from the 2021 Canadian census [@governmentofcanadaCensusPopulation2023], the Open Data Hamilton Portal [@opendatahamiltonCityBoundary2023] and the USGS [@greatlakesUSGS2010]. Scale is in quartiles."
knitr::include_graphics(glue::glue(here::here(), "/figures/percats_copp_plot.png"))
```

```{r plot-figure5, out.width=450, out.height=450}
#| label: fig-Fig5
#| fig-cap: "The diversity measure based on the proportion of care category accessibility in Figure 4. These values are also a summary of the 15th input into the SOM. Basemap shapefiles are retrieved from the 2021 Canadian census [@governmentofcanadaCensusPopulation2023], the Open Data Hamilton Portal [@opendatahamiltonCityBoundary2023] and the USGS [@greatlakesUSGS2010]. Scale is in quartiles."
knitr::include_graphics(glue::glue(here::here(), "/figures/all_cats_entropy_copp_plot.png"))
```

@@

In @fig-Fig4, for all care categories, access to opportunities is concentrated in downtown Hamilton (Hamilton Central), with the highest concentration (top quantiles) near the lake shore of Hamilton Central. Grocery-centric destinations follow this trend most prominently, followed by health-centric and dependent-centric caring destinations. Access to errand-centric destinations appears to most spatially homogeneous of the four care destination types, though the top quantiles are still concentrated in Hamilton Central. As the 15-minute walking accessibility to destinations is a spatially weighted summary measure of the count of destinations, @fig-Fig4 mirrors the spatial distribution in @fig-Fig3. In @fig-Fig5, areas that have high diversity are broadly located in downtown Hamilton as well, however, there are exceptions. Namely, there are pockets of high diversity outside the downtown core that that score low levels of accessibility for all care categories. Similarly, there are areas with low diversity within the downtown core that have only moderate or high accessibility to certain care destination types.

Based on the SOM methodology discussed and using inputs summarised in @fig-Fig4 and @fig-Fig5, seven superclusters are identified as shown in the dissimilarity-index-based dendrogram (left plot of @fig-Fig6). Seven superclusters are selected to represent the data, as they explain a high amount of the variance (over 90%; right plot in @fig-Fig6). Each supercluster can be intuitively labelled based on the quantity of accessibility and diversity in accessibility per care category offered by a parcel, as mapped onto the SOM input grid. Labels representing grades A through D are assigned to qualify the seven superclusters. Higher grades (A+ and A) corresponding to the highest accessibility and diversity, while lowest grades (D) represent the lowest accessibility and diversity scores.

```{r plot-figure6, out.width = '100%', out.height = '100%'}
#| label: fig-Fig6
#| fig-cap: "The seven resulting superclusters from the SOM output are represented in the dendrogram (left). As a diagnostic, the proportion of variance unexplained by the number superclusters is displayed on the right, with red lines indicating seven superclusters."
knitr::include_graphics(glue::glue(here::here(), "/figures/all_access_sc_plot_small.png"))
```

More specifically, each supercluster was labelled a grade by the authors by referring to the supercluster's descriptive statistics, as summarised in @tbl-Tbl2 and the boxplots in @fig-Fig7. Additionally, the grades for each supercluster are further explained in context of the proportion of urban, suburban and rural parcels as classified by the City of Hamilton within the parcel dataset [@Teranet_2021]. Notably, the majority of parcels in Hamilton are considered urban (`r ((access_15_urban_summary |> filter(URBAN_RURA == "U") |> select(count_urb_parcels) |> sum()) / (access_15_urban_summary  |> select(count_urb_parcels) |> sum() )) |> scales::percent(0.1)`), while suburban and rural parcels constitute the remaining `r ((access_15_urban_summary |> filter(URBAN_RURA == "S") |> select(count_urb_parcels) |> sum()) / (access_15_urban_summary  |> select(count_urb_parcels) |> sum() )) |> scales::percent(0.1)` and `r ((access_15_urban_summary |> filter(URBAN_RURA == "R") |> select(count_urb_parcels) |> sum()) / (access_15_urban_summary  |> select(count_urb_parcels) |> sum() )) |> scales::percent(0.1)`, respectively:

<!--"(A+) Exceptional: very caring & very complete", #C1
"(A ) Excellent: very caring & very complete ", #(C2)
"(A-) Dep- heavy, otherwise excellent: very caring & not complete", #(C3)

"(B+) Above average: caring & very complete", # (C6)
"(B ) Average: caring & complete", #(C5)
"(B- ) Dep- heavy, otherwise below average: somewhat caring & not complete", # (C4)

"(D ) Below average: not caring & not complete") # (C7)-->

- A+ and A represent superclusters with exceptionally high caring accessibility for all destination types and high diversity scores in the top quantile or above. Together, these grades represent 24% of all parcels in the city and `r ((access_15_urban_summary |> filter((scluster == 1 | scluster == 2) & (URBAN_RURA == "U")) |> select(count_urb_parcels) |> sum()) / (access_15_urban_summary |> filter((scluster == 1 | scluster == 2)) |> select(count_urb_parcels) |> sum() )) |> scales::percent()` are located in urban neighbourhoods. Based on the spatial access and diversity of care destination access offered by these neighbourhoods, they could be called "very caring" and "very complete".
- A- has high caring accessibility, but low diversity. These neighbourhoods could be classified as "very caring" but and "not complete". Interestingly, A- is like A but with much higher dependent-centric destinations, particularly parks, schools, and daycares, with only moderately high scores for all other destination types. This disbalance contributes to reducing its overall diversity score. However, it could be characterized as a supercluster that demonstrates potential in being retrofitted to provide A+ or A level of completely caring access. 4% of all parcels are represented by the A- grade and, similar to parcels in A+ and A superclusters, `r ((access_15_urban_summary |> filter((scluster == 3) & (URBAN_RURA == "U")) |> select(count_urb_parcels) |> sum()) / (access_15_urban_summary |> filter((scluster == 3)) |> select(count_urb_parcels) |> sum() )) |> scales::percent()` are located in urban neighbourhoods. 
- B+ and B superclusters present about average "completely caring" access. These superclusters can serve as the benchmark for what 'average' completely caring access in Hamilton currently looks like. These grades represent 15% of all parcels and like parcels in A+, A and A- superclusters, `r ((access_15_urban_summary |> filter((scluster == 6 | scluster == 5) & (URBAN_RURA == "U")) |> select(count_urb_parcels) |> sum()) / (access_15_urban_summary |> filter((scluster == 6 | scluster == 5)) |> select(count_urb_parcels) |> sum() )) |> scales::percent()` are located in urban neighbourhoods.
- B- provides above average dependent-centric destination access, particularly parks, daycares and schools, but below average access to other destination types and hence has low diversity scores. B- neighbourhoods could be labelled "somewhat caring" but "not complete". B- is like A- as it demonstrates complete caring 15-minute potential if retrofitted. As these parcels demonstrate caring access to some destinations, they may have the potential to be retrofitted to support complete access to all caring destination types. This grade represents 19% of all parcels. While the majority of parcels are still classified as urban neighbourhoods (`r ((access_15_urban_summary |> filter((scluster == 4) & (URBAN_RURA == "U")) |> select(count_urb_parcels) |> sum()) / (access_15_urban_summary |> filter((scluster == 4)) |> select(count_urb_parcels) |> sum() )) |> scales::percent()`), this supercluster represents `r ( (access_15_urban_summary |> filter((scluster == 4) & (URBAN_RURA == "S")) |> select(count_urb_parcels) |> sum()) / (access_15_urban_summary |> filter(URBAN_RURA == "S") |> select(count_urb_parcels) |> sum()) ) |> scales::percent(0.1)` of the suburban parcels and `r ( (access_15_urban_summary |> filter((scluster == 4) & (URBAN_RURA == "R")) |> select(count_urb_parcels) |> sum()) / (access_15_urban_summary |> filter(URBAN_RURA == "R") |> select(count_urb_parcels) |> sum()) ) |> scales::percent(0.1)` rural parcels in the city. These suburban and rural parcels are in proximity to urban parcels, as elaborated in the next subsection. 
- D superclusters demonstrate the lowest scores all-around, representing room for land-use improvement that addresses complete and caring 15-minute access. D neighbourhoods could be classified as "not caring" and "not complete" due to the low amount of access to all care destinations offered and the care category diversity of this access. Notably, this supercluster characterizes the largest number of parcels, representing 38% of parcels in the city. Only `r ((access_15_urban_summary |> filter((scluster == 7) & (URBAN_RURA == "U")) |> select(count_urb_parcels) |> sum()) / (access_15_urban_summary |> filter((scluster == 7)) |> select(count_urb_parcels) |> sum() )) |> scales::percent()` of these parcels are 'urban', and this supercluster represents the vast majority of suburban and rural parcels in the city (`r ( (access_15_urban_summary |> filter((scluster == 7) & (URBAN_RURA == "S")) |> select(count_urb_parcels) |> sum()) / (access_15_urban_summary |> filter(URBAN_RURA == "S") |> select(count_urb_parcels) |> sum()) ) |> scales::percent(0.1)` and `r ( (access_15_urban_summary |> filter((scluster == 7) & (URBAN_RURA == "R")) |> select(count_urb_parcels) |> sum()) / (access_15_urban_summary |> filter(URBAN_RURA == "R") |> select(count_urb_parcels) |> sum()) ) |> scales::percent(0.1)`).

@@

To spatially demonstrate where the superclusters are located, representative supercluster grades for each DA are visualised in @fig-Fig8. This visualisation is created by grouping parcels by their DA and selecting the grade that is most dominant within that DA. For reference, the median number of parcels in each DA is `r df_DA$parcel_count |> median() |> round()` and supercluster grade membership within a DA is typically pure, with the median membership being `r df_DA$max_label_val |> median() |> scales::percent()` of a single cluster.

```{r plot-figure8, out.width=800, out.height=450}
#| label: fig-Fig8
#| fig-cap: "The maximum median parcel supercluster membership per DA. Escarpment is visualised as a grey dashed line. Basemap shapefiles are retrieved from the 2021 Canadian census [@governmentofcanadaCensusPopulation2023], the Open Data Hamilton Portal [@opendatahamiltonCityBoundary2023] and the USGS [@greatlakesUSGS2010]."
knitr::include_graphics(glue::glue(here::here(), "/figures/SClabel_plot.png"))
```

Recall, the superclusters in @fig-Fig8 are a combination of the cumulative opportunity and diversity of access measures translated into meaningful typologies through the SOM methodology. To reiterate, accessibility is the measure of _potential_ interaction, e.g., how many care destinations one could reach within a 15-minute walk, these typologies are useful in identifying which areas of the city are providing city-wide high levels of complete caring access, which areas are city-wide average, and which areas are city-wide below average.

In @fig-Fig8, it can be observed that the excellent and exceptional neighbourhoods (A, A+), which are very caring and very complete 15-minute neighbourhoods, are located within the center of the city and closest to the shoreline in the downtown core of Hamilton. The above and below average neighbourhoods (B+, B) are often proximate to the excellent and exceptional neighbourhoods and are more prevalent in south of the escarpment within the center Hamilton. The escarpment is a physical barrier, with few pedestrian-accessible access points to traverse; hence the typologies describing neighbouring DAs separated by the escarpment are often different. Below average (D) neighbourhoods are located in peripheral areas outside the center of Hamilton, in areas where urban form is characterised by lower density residential housing and auto-dependent mobility (majority rural and suburban). D neighbourhoods also tend to have the lowest diversity scores, with either no access or access to only one type of care destination category.
